---
title: "Getting started with endogCopula"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with endogCopula}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Copula-based endogeneity correction

Endogeneity arises whenever a regressor is correlated with the error term of
a regression model, for example because of omitted variables, simultaneity,
or measurement error. Ordinary least squares is then biased and
inconsistent. The classical remedy is an instrumental variable, but valid
and strong instruments are notoriously hard to find. Park and Gupta (2012)
proposed an instrument-free alternative: model the joint distribution of the
endogenous regressor and the structural error with a Gaussian copula. Under
the assumption that the endogenous regressor is non-normally distributed
while the error is normal, the dependence between the two can be recovered
from the data itself, and adding a "copula control function" -- a
transformation of the endogenous regressor's estimated cumulative
distribution function (CDF) -- to the regression removes the endogeneity
bias.

A lively methodological literature has refined this idea. Becker, Proksch,
and Ringle (2022) revisited the original approach and highlighted the
conditions under which it works well, while Qian, Koschmann, and Xie (2024)
provide a practical guide. Newer estimators address correlation between
endogenous and exogenous regressors, which the original approach ignores:
Yang, Qian, and Xie (2025) introduce a two-stage copula generated regressor
approach (2sCOPE), Hu, Qian, and Xie (2025) develop a nonparametric
two-stage variant (2sCOPE-np), Breitung, Mayer, and Wied (2024) study
corrections based on nonlinear transformations, Haschka (2024) exploits
between-regressor correlation, and Liengaard et al. (2025) propose an
adjusted estimator with improved finite-sample behaviour.

The `endogCopula` package collects these cross-sectional estimators behind a
single formula interface and provides bootstrap standard errors plus
`print()`, `summary()`, `coef()`, `residuals()`, and `confint()` methods.
The implementations are tested for numerical equivalence against the
published replication code accompanying the papers.

## Estimators and references

| Function            | Estimator  | Reference                          |
|---------------------|------------|------------------------------------|
| `CopRegPG()`        | PG         | Park and Gupta (2012)              |
| `CopReg2sCOPE()`    | 2sCOPE     | Yang, Qian, and Xie (2025)         |
| `CopReg2sCOPEnp()`  | 2sCOPE-np  | Hu, Qian, and Xie (2025)           |
| `CopRegIMA()`       | IMA        | Haschka (2024)                     |
| `CopRegBMW()`       | BMW        | Breitung, Mayer, and Wied (2024)   |
| `CopRegJAMS()`      | JAMS       | Liengaard et al. (2025)            |
| `CopRegML_par()`    | PANEL      | Haschka (2022), via `endogCopulaPanel` |
| `CopRegBayes()`     | BAYES      | Haschka (2022b), via `endogCopulaBayes` |

All cross-sectional estimators use a two-part formula,

```r
y ~ endog1 + endog2 | exog1 + exog2
```

where the variables before the bar are the continuous endogenous regressors
and the variables after the bar are exogenous. Add `- 1` to drop the
intercept and wrap dummies in `as.factor()`.

## A worked example

The package ships a simulated dataset, `sim_endog`, generated from

$$y = 1 + 2\, z_{\text{endog}} + 1.5\, x_{\text{exog}} + u,$$

where a latent Gaussian factor makes `z_endog` correlated with the
structural error `u`, and `w_instr` shifts `z_endog` without entering the
outcome equation.

```{r setup}
library(endogCopula)
data(sim_endog)
str(sim_endog)
```

A naive OLS regression that ignores the endogeneity of `z_endog` produces a
biased estimate of its coefficient (true value: 2):

```{r ols}
ols <- lm(y ~ z_endog + x_exog, data = sim_endog)
coef(ols)
```

The Park and Gupta (2012) estimator adds a copula control function for
`z_endog`. We use `nboots = 25` here so the vignette builds quickly; in
practice several hundred bootstrap replicates are recommended.

```{r pg, message = FALSE}
set.seed(42)
fit_pg <- CopRegPG(
  y ~ z_endog | x_exog + w_instr,
  data = sim_endog,
  cdf = "resc.ecdf",
  nboots = 25
)
summary(fit_pg)
```

The 2sCOPE estimator of Yang, Qian, and Xie (2025) additionally accounts
for correlation between the endogenous and the exogenous regressors:

```{r scope, message = FALSE}
set.seed(42)
fit_2scope <- CopReg2sCOPE(
  y ~ z_endog | x_exog + w_instr,
  data = sim_endog,
  cdf = "resc.ecdf",
  nboots = 25
)
summary(fit_2scope)
```

Comparing the estimates of the endogenous coefficient:

```{r compare}
c(
  OLS      = unname(coef(ols)["z_endog"]),
  PG       = unname(coef(fit_pg)["z_endog"]),
  `2sCOPE` = unname(coef(fit_2scope)["z_endog"])
)
```

Confidence intervals based on the bootstrap replicates are available via
`confint()`:

```{r confint}
confint(fit_pg, parm = "z_endog")
confint(fit_pg, parm = "z_endog", type = "percentile")
```

## Choosing the `cdf` argument

The copula control function depends on an estimate of each endogenous
regressor's marginal CDF, selected via the `cdf` argument:

- `"kde"` -- the integral of a kernel density estimator, as used by Park
  and Gupta (2012) and Haschka (2022). Requires the suggested `ks` package.
- `"ecdf"` -- the empirical CDF with its boundary values replaced, proposed
  by Becker, Proksch, and Ringle (2022).
- `"resc.ecdf"` -- a rescaled empirical CDF proposed by Qian, Koschmann,
  and Xie (2024) and used by Haschka (2024) and Yang, Qian, and Xie (2025).
- `"adj.ecdf"` -- an adjusted empirical CDF proposed by Liengaard et al.
  (2025).

A pragmatic default is to use the CDF estimator from the paper that
introduced the estimator you are applying (e.g. `"adj.ecdf"` with
`CopRegJAMS()`, `"resc.ecdf"` with `CopReg2sCOPE()`), and to check that the
conclusions are robust across the alternatives.

## Panel and Bayesian estimators

The fixed-effects panel estimator of Haschka (2022) and the Bayesian
sampler of Haschka (2022b) live in dedicated companion packages:

- [`endogCopulaPanel`](https://github.com/ashgreat/endogCopulaPanel) for
  `CopRegML_par()`, the maximum likelihood panel estimator.
- [`endogCopulaBayes`](https://github.com/ashgreat/endogCopulaBayes) for
  `CopRegBayes()`, the MCMC sampler.

When these packages are installed, the wrappers of the same names exported
by `endogCopula` delegate to them.

## References

- Becker, J.-M., D. Proksch, and C. M. Ringle (2022). Revisiting Gaussian
  copulas to handle endogenous regressors. *Journal of the Academy of
  Marketing Science* 50, 46--66.
- Breitung, J., A. Mayer, and D. Wied (2024). Asymptotic properties of
  endogeneity corrections using nonlinear transformations. *The
  Econometrics Journal* 27(3), 362--383.
- Haschka, R. E. (2022). Handling endogenous regressors using copulas: A
  generalisation to linear panel models with fixed effects and correlated
  regressors. *Journal of Marketing Research* 59(4), 860--881.
- Haschka, R. E. (2022b). Bayesian inference for joint estimation models
  using copulas to handle endogenous regressors.
  <https://ssrn.com/abstract=4235194>
- Haschka, R. E. (2024). Robustness of copula-correction models in causal
  analysis: Exploiting between-regressor correlation. *IMA Journal of
  Management Mathematics* 36(1), 161--180.
- Hu, X., Y. Qian, and H. Xie (2025). Correcting endogeneity via
  instrument-free two-stage nonparametric copula control functions. NBER
  Working Paper 33607. <http://www.nber.org/papers/w33607>
- Liengaard, B. D., J.-M. Becker, M. Bennedsen, P. Heiler, L. N. Taylor,
  and C. M. Ringle (2025). Dealing with regression models' endogeneity by
  means of an adjusted estimator for the Gaussian copula approach. *Journal
  of the Academy of Marketing Science* 53, 279--299.
- Park, S. and S. Gupta (2012). Handling endogenous regressors by joint
  estimation using copulas. *Marketing Science* 31(4), 567--586.
- Qian, Y., A. Koschmann, and H. Xie (2024). A practical guide to
  endogeneity correction using copulas. NBER Working Paper 32231.
  <https://www.nber.org/papers/w32231>
- Yang, F., Y. Qian, and H. Xie (2025). Addressing endogeneity using a
  two-stage copula generated regressor approach. *Journal of Marketing
  Research* 62(4), 601--623.
